---
title: "Ping River Streamflow Reconstruction with Linear Dynamical System"
author: "Hung Nguyen"
date: "2020-04-10"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float: yes
    highlight: haddock
---

# Introduction

This repository reproduces the results by Nguyen and Galelli (2018), which reconstructs streamflow for the Ping River at station P.1. This work explores a novel reconstruction method using a Linear Dynamical System (LDS) model instead of the more conventionally used linear regression models. An LDS model is formulated as follows (Shumway and Stoffer, 1982; Ghahramani and Hinton, 1996).

$$
\begin{align}
  x_{t+1} &= Ax_t + Bu_t + \eta_t\\
  y_t &= Cx_t + Dv_t + \varepsilon_t \\
  \eta_t & \sim \mathcal{N}(0, Q)\\
  \varepsilon_t &\sim \mathcal{N}(0, R)\\
  x_1 &\sim \mathcal{N}(0, V_1)
\end{align}
$$

In the context of streamflow reconstruction, we can take $v = u$, which leads to the model presented in the paper.

The LDS model parameters, together with the hidden state trajector $x$, are learned with an Expectation-Maximization (EM) algorithm. The details of this algorithm are presented in the paper, and the algrorithm has been implemented in the R package `ldsr`, adapted from the MATLAB code of Cheng and Sabes (2006). We will now load `ldsr` together with several packages that are used frequently in this repository. Other packages will be referred to with `::` when needed.

```{r setup, warning=FALSE, message=FALSE}
library(here)        # Pathing
library(data.table)  # Data wrangling
library(ldsr)        # Reconstruction
library(ggplot2)     # Plotting
library(patchwork)   # Arranging plots
```

# Data

* Streamflow data: annual streamflow at station P.1, aggregated from monthly streamflow according to the hydrological year (April-March). Monthly streamflow was retrieved from the Thai Royal Irrigation Department.
* Proxy data: 
    - The Monsoon Asia Drought Atlas (MADA) (Cook *et al*, 2010).
    - Tree ring data from the [Dendrobox project.](http://dendrobox.org)

```{r read data}
# STREAMFLOW DATA
Qmonth <- fread(here('data/P1-monthly.csv'))
Qmonth <- melt(Qmonth, id.var = 'year', variable.name = 'month', value.name = 'Q')
Qann <- fread(here('data/P1-annual.csv'))
P1_xy <- c(99.00443, 18.78766) # P.1 coordinates
P1_df <- data.frame(long = P1_xy[1], lat = P1_xy[2]) # For plotting

# PALEO DATA
##  MADA
madaMeta <- fread(here('data/mada_xy.csv'), colClasses = c(ID = 'character'))
mada <- fread(here('data/jja-mada.txt'))[year >= 1600]
madaMat <- as.matrix(mada[, -'year'])
## Tree ring
crnMeta <- fread(here('data/tree-ring/crn-meta.csv'))
crn <- crnMeta[, fread(here(paste0('data/tree-ring/', .BY, '.csv'))), by = ID]
allSites <- fread(here('data/tree-ring/all-chronology-locations.csv'))

# BACKGROUND MAP
countryBoundaries <- sf::read_sf(here('data/geo/mada-region-country-boundaries.gpkg'))
```

Streamflow seasonality (Figure 1b).

```{r, fig.width=6, fig.height=3}
ggplot(Qmonth) +
  geom_boxplot(aes(month, Q)) +
  theme_classic() +
  labs(x = NULL, y = 'Monthly streamflow [million m\u00B3]')
```

# Methodology

## Input selection

### MADA grid points

Calculate the distance between each MADA grid point and P.1.

```{r}
madaMeta[, d_P1 := geosphere::distGeo(c(long, lat), P1_xy) / 1000, by = ID]
```

Find the MADA grid points that are within 1,200 km from P.1, and calculate their correlations with P.1 streamflow.

```{r}
selectedMada <- madaMeta[d_P1 <= 1200][, cor_P1 := cor(madaMat[322:406, ID], Qann$Qa)]
```

Plot the study region on the map (Figure 2).

```{r, fig.width=6, fig.height=5}
extend <- c(-1, 1)
longRange <- selectedMada[, range(long) + extend]
latRange <- selectedMada[, range(lat) + extend]
ggplot() +
  geom_sf(data = countryBoundaries, fill = NA) +
  geom_point(aes(long, lat), 
             data = allSites[long %between% longRange & lat %between% latRange], 
             shape = 17, colour = 'darkgreen', size = 2.5) +
  geom_point(aes(long, lat, colour = cor_P1), data = selectedMada, size = 4) +
  geom_point(aes(long, lat), data = P1_df, colour = 'darkred', size = 2, shape = 15) +
  geom_text(aes(long, lat), label = 'P.1', data = P1_df, 
            nudge_x = 0.5, nudge_y = -0.5, size = 4) +
  scale_x_continuous(expand = c(0, 0), limits = longRange) +
  scale_y_continuous(expand = c(0, 0), limits = latRange) +
  scale_colour_gradient2(name = 'Correlation', low = 'salmon', high = 'steelblue') +
  labs(x = NULL, y = NULL) +
  coord_sf() +
  theme_bw() +
  theme(legend.key.height = unit(1, 'cm'))
```

Compare MADA-streamflow correlations with chronology-streamflow correlations (Table 1).

```{r, rows.print = 11}
crnMeta[, d_P1 := geosphere::distGeo(c(long, lat), P1_xy) / 1000, by = ID]
crnCor <- crn[Qann, on = 'year'
            ][, 
              { 
                out <- cor.test(rwi, Qa)
                list(cor_P1 = round(out$estimate, 2),
                     p.value = round(out$p.value, 4))
                }, 
              by = ID
            ][crnMeta, on = 'ID'
            ][, -'name']
madaCor <- mada[322:406, c('year', 251, 252, 275, 301)
              ][, melt(.SD, id.var = 'year', variable.name = 'ID', value.name = 'pdsi')
              ][Qann, on = 'year'
              ][, 
                { 
                  out <- cor.test(pdsi, Qa)
                  list(cor_P1 = round(out$estimate, 2),
                       p.value = round(out$p.value, 4))
                }, 
                by = ID
              ][madaMeta, on = 'ID', nomatch = NULL]
corCompare <- rbind(crnCor, madaCor)
corCompare[order(d_P1)][, .(ID, d_P1 = round(d_P1, 0), cor_P1, p.value)]
```

### Principal Component Analysis

```{r}
pca <-  summary(prcomp(madaMat[, selectedMada$ID], scale. = TRUE))
pca
```

The first 12 PCs account for 95% of the variance. 

### Correlations between PCs and streamflow

Now, we perform a bootstrapped correlation analysis on these PCs using a utility function called `cor_pivot()`, which calculates the correlation between one column (the pivot column) of a matrix and each of its other columns.

```{r}
cor_pivot <- function(X, col = ncol(X)) cor(X[, -col], X[, col])
```

```{r}
# Extract the instrumental period
mat <- cbind(pca$x[322:406, 1:12], log(Qann$Qa))
bootCor <- boot::tsboot(mat, cor_pivot, 1000, sim = 'geom', l = 50)
bootCor
```

We determine if the correlations are significant by checking the 5%-95% quantile range.

```{r}
corQuantiles <- apply(bootCor$t, 2, function(x) quantile(x, probs = c(0.05, 0.5, 0.95)))
corQuantiles
```

From this result, we see that PCs 1, 3, 4, 6, 9, 10, 11, and 12 are significantly correlated with streamflow. We can also visualize these correlations.

```{r, fig.width=7, fig.height=3}
corDT <- as.data.table(t(corQuantiles))
setnames(corDT, c('low', 'median', 'high'))
corDT$PC <- paste0('PC', 1:12)
corDT[, PC := factor(PC, levels = PC)]
corDT[, signif := !between(0, low, high)]
ggplot(corDT) +
  geom_linerange(aes(PC, ymin = low, ymax = high, colour = signif), size = 1) +
  geom_point(aes(PC, median, colour = signif), size = 2) +
  geom_hline(yintercept = 0, colour = 'gray') +
  scale_colour_manual(name = 'Significance', values = c('lightblue', 'steelblue')) +
  labs(x = NULL, y = 'Correlation') + 
  theme_classic()
```

### Backward selection

Next, we select the principal components using a backward procedure based on the Bayesian Information Criterion (BIC).

```{r}
X <- pca$x[, 1:12]
ivs <- summary(leaps::regsubsets(X[322:406, ], log(Qann$Qa), method = 'backward'))
ivs$bic
```

The fifth model has the lowest BIC.

```{r}
names(which(ivs$which[which.min(ivs$bic), ]))[-1]
```

The selected PCs are 1, 3, 6, 9, and 12. All the selected PCs are significantly correlated with streamflow, so this is a reasonable selection.

## Reconstruction

### Linear regression

We build and cross-validate a linear regression reconstruction to use as a benchmark for our LDS model.

```{r}
# Input
pc <- data.table(X[, c(1, 3, 6, 9, 12)])
bm <- PCR_reconstruction(Qann, pc, start.year = 1600)
# Cross-validate
set.seed(321) # To make the cross-validation folds reproducible
Z <- replicate(100, sort(sample(1:85, 9)), simplify = FALSE)
bmcv <- cvPCR(Qann, pc, start.year = 1600, Z = Z, metric.space = 'original')
```

### Linear dynamical system

Since EM is a local search routine, we run it with multiple restarts, each of which has a different initial condition, i.e., different values for $\theta_0$. These different restarts can be run in parallel to speed up the computation. The `ldsr` packages implements parallelism with the `foreach` package. Following the `foreach` philosphy, it is up to the user to choose the parallel backend. Here, we use the `doFuture` backend.

```{r, eval=FALSE}
doFuture::registerDoFuture()
future::plan(future::multiprocess)
```

From our experience, about 20-50 restarts is suffient to learn a good model. The `ldsr` package takes 50 restarts by default. For cross-validation, we can reduce the number of restarts to 20 to speed up the cross-validation. On a 4-core 3.4 GHz desktop, learning takes about 2-3 seconds, and cross-validation takes about a minute.

```{r parallel, eval=FALSE, include=FALSE}
# This chunk does the actual reconstruction. It takes several minutes on a 4-core 3.4 GHz desktop. It is not run during vignette production to save time.
doFuture::registerDoFuture()
future::plan(future::multiprocess)
u <- v <- t(pc)
# Build model
lds <- LDS_reconstruction(Qann, u, v, start.year = 1600)
ldscv <- cvLDS(Qann, u, v, start.year = 1600, Z = Z, num.restarts = 20, metric.space = 'original')
saveRDS(lds, here('results/lds.RDS'))
saveRDS(ldscv, here('results/ldscv.RDS'))
``` 

```{r, include=FALSE}
# This chunk reads the pre-produced results so that the vignette can be built quickly.
u <- v <- t(pc)
lds <- readRDS(here('results/lds.RDS'))
ldscv <- readRDS(here('results/ldscv.RDS'))
```

```{r, eval=FALSE}
# Take u and v to be the same matrix, which is the selected PCs.
# Because of matrix convention used in the algorithm, the PC matrix has to be transposed.
u <- v <- t(pc)
# Build model
lds <- LDS_reconstruction(Qann, u, v, start.year = 1600)
# Cross-validate the LDS model using the same folds as with linear regression.
ldscv <- cvLDS(Qann, u, v, start.year = 1600, num.restarts = 20, Z = Z, 
               metric.space = 'original')
```

# Results

## Model parameters

```{r}
lds$theta
```

Compare D with linear regression coefficients (Table 2)

```{r}
cbind(beta = bm$coeffs[-1], D = c(lds$theta$D))
```

## Performance comparison

Figure 4.

```{r, fig.width=5, fig.height=5}
scores <- rbindlist(list(LR = bmcv$metrics.dist, LDS = ldscv$metrics.dist),
                    idcol = 'model') 
scores <- melt(scores, id.vars = 'model', 
       measure.vars = c('R2', 'RE', 'CE', 'nRMSE'), 
       variable.name = 'metric')

scores[metric == 'R2', metric := 'R\u00B2']
scores[, metric := factor(metric, levels = c('R\u00b2', 'RE', 'CE', 'nRMSE'))]

ggplot(scores, aes(model, value)) +
  geom_jitter(colour = 'gray70', width = 0.2) +
  stat_summary(geom = 'point', fun = 'mean', colour = 'red') +
  facet_wrap(vars(metric), scales = 'free') +
  labs(x = NULL, y = 'Metric value') +
  theme_classic() +
  theme(strip.background = element_blank())
```

Note: in the original version of the paper, the performance of LDS was higher than that of linear regression. This turned out to be because of an error in the cross-validation procedure. We have corrected that; LDS and linear regression has equal performance. But LDS gives you a state trajectory as an added benefit.

## Instrumental period

Figure 5.

```{r, fig.width=8, fig.height=4}
p1 <- ggplot(lds$rec[year %in% Qann$year]) +
  geom_ribbon(aes(year, ymax = Ql, ymin = Qu), fill = 'gray90') +
  geom_line(aes(year, Q, colour = 'LDS', linetype = 'LDS')) +
  geom_line(aes(year, Qa, colour = 'Inst', linetype = 'Inst'), data = Qann) +
  geom_line(aes(year, Q, colour = 'LR', linetype = 'LR'), 
            data = bm$rec[year %in% Qann$year]) +
  scale_linetype_manual(name = NULL, values = c(1, 1, 2)) +
  scale_colour_manual(name = NULL, values = c('darkorange', 'black', 'black')) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = 'Annual streamflow [million m\u00B3]') +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- ggplot(lds$rec[year %in% Qann$year]) +
  geom_ribbon(aes(year, ymax = Xl, ymin = Xu), fill = 'gray90') +
  geom_hline(yintercept = 0, colour = 'gray50') +
  geom_line(aes(year, X)) +
  labs(x = NULL, y = 'Catchment state [-]') +
  theme_classic()

p1 / p2 + plot_layout(heights = c(1, 0.6))
```

## Residuals

Figure 6.

```{r, fig.width=6, fig.height=6}
bmr <- log(bm$rec$Q[322:406]) - log(Qann$Qa)
ldsr <- log(lds$rec$Q[322:406]) - log(Qann$Qa)

probs = c(0.1, 0.9)
par(mfrow = c(2,2), ps = 10)

par(mai = c(0.7, 0.7, 0.6, 0.2))

qqnorm(bmr, ylim = c(-1, 1), main = '', lwd = 1, pch = 16)
qqline(bmr, probs = probs, qtype = 5)
mtext('Linear regression', 3, line = 2.5, font = 2)
mtext('(a)', 3, line = 1, adj = 0, font = 2, cex = 0.75)

qqnorm(ldsr, ylim = c(-1, 1), main = '', lwd = 1, pch = 16)
qqline(ldsr, probs = probs, qtype = 5)
mtext('Linear dynamical system', 3, line = 2.5, font = 2)
mtext('(b)', 3, line = 1, adj = 0, font = 2, cex = 0.75)

par(mai = c(0.7, 0.7, 0.3, 0.2))
acf(bmr, main = '')
mtext('(c)', 3, line = 1, adj = 0, font = 2, cex = 0.75)

acf(ldsr, main = '')
mtext('(d)', 3, line = 1, adj = 0, font = 2, cex = 0.75)
```

## History of regime shifts

Figure 7.

```{r, fig.width=8, fig.height=4}
# Drought periods
modelDroughts <- data.frame(startYear = c(1638, 1672, 1791, 1877), 
                            endYear = c(1658, 1695, 1797, 1879))
megadroughts <- data.frame(startYear = c(1638, 1756, 1789.5, 1792, 1876),
                           endYear = c(1641, 1768, 1790.5, 1796, 1878))

p1 <- ggplot(lds$rec) +
  geom_ribbon(aes(year, ymax = Ql, ymin = Qu), fill = 'gray90') +
  geom_hline(aes(yintercept = mean(Q)), colour = 'red2') +
  geom_line(aes(year, Q, colour = 'LR', linetype = 'LR'), data = bm$rec) +
  geom_line(aes(year, Q, colour = 'LDS', linetype = 'LDS')) +
  scale_colour_manual(name = NULL, values = c('black', 'steelblue')) +
  scale_linetype_manual(name = NULL, values = c(1, 2)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = 'Annual streamflow [million m\u00B3]') +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())

p2 <- ggplot(lds$rec) +
  geom_ribbon(aes(year, ymax = Xl, ymin = Xu), fill = 'gray90') +
  geom_hline(yintercept = 0, colour = 'red2') +
  geom_line(aes(year, X)) +
  geom_rect(aes(xmin = startYear, xmax = endYear, ymin = -Inf, ymax = Inf), 
            data = modelDroughts, fill = 'yellow', alpha = 0.4) +
  geom_rect(aes(xmin = startYear, xmax = endYear, ymin = -Inf, ymax = Inf), 
            data = megadroughts, fill = 'darkorange', alpha = 0.4) +
  labs(x = NULL, y = 'Catchment state [-]') +
  theme_classic()

p1 / p2 + plot_layout(heights = c(1, 0.6))
```

## Wavelet analysis

We compare the wavelet spectrum of the reconstructed streamflow with that of the reconstructed SST by Tierney et al (2015).

```{r, results='hide', fig.keep=TRUE, fig.width=6, fig.height=6}
# Wavet analysis of reconstructed streamflow
lds.w <- WaveletComp::analyze.wavelet(lds$rec, 'Q', loess.span = 0, verbose = FALSE)
lds.w$axis.1 <- lds.w$axis.1 + 1600
lds.w$coi.1 <- lds.w$coi.1 + 1600
# Read SST reconstruction
epacSST <- fread(here('data/epac-sst.csv'))
# Wavelet analysis of SST reconstruction
sst.w <- WaveletComp::analyze.wavelet(epacSST, 'sst', loess.span = 0, verbose = FALSE)
sst.w$axis.1 <- sst.w$axis.1 + 1607
sst.w$coi.1 <- sst.w$coi.1 + 1607
## Plots
par(mfrow = c(2,1), mar = c(3, 4, 1, 2))
lds.wimage <- WaveletComp::wt.image(lds.w, 
                                    timelab = 'Year', 
                                    periodlab = 'Period', 
                                    color.key = 'q', 
                                    plot.ridge = FALSE, 
                                    plot.contour = FALSE,
                                    graphics.reset = FALSE)
sst.wimage <- WaveletComp::wt.image(sst.w, 
                                    timelab = 'Year',
                                    periodlab = 'Period', 
                                    color.key = 'q', 
                                    plot.ridge = FALSE, 
                                    plot.contour = FALSE)
```

## Stochastic replicates

Build 100 stochastic replicates.

```{r}
set.seed(500)
reps <- LDS_rep(lds$theta, u, v, years = lds$rec$year, mu = mean(log(Qann$Qa)))
```

Plot the replicates (Figure 9).

```{r}
plot_replicates <- function(reps, rec) {

  p1 <- ggplot(reps) +
    geom_line(aes(year, simQ, group = rep), colour = 'gray90') +
    geom_line(aes(year, Q), data = rec, colour = 'black') +
    labs(x = NULL, y = 'Annual streamflow [million m\u00B3]') +
    theme_classic() +
    theme(axis.line.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
  
  p2 <- ggplot(reps) +
    geom_line(aes(year, simX, group = rep), colour = 'gray90') +
    geom_line(aes(year, X), data = rec, colour = 'black') +
    labs(x = 'Year',
         y = 'Flow regime') +
    theme_classic()
  
  p1 / p2 + plot_layout(heights = c(1, 0.6))
}
```

```{r, fig.width=7, fig.height=4}
plot_replicates(reps, lds$rec)
```

# References

Cook, E. R., Anchukaitis, K. J., Buckley, B. M., D’Arrigo, R. D., Jacoby, G. C., & Wright, W. E. (2010). Asian Monsoon Failure and Megadrought During the Last Millennium. Science, 328(5977), 486–489. https://doi.org/10.1126/science.1185188

Cheng, S., & Sabes, P. N. (2006). Modeling Sensorimotor Learning with Linear Dynamical Systems. Neural Computation, 18(4), 760–793. https://doi.org/10.1162/089976606775774651

Ghahramani, Z., & Hinton, G. E. (1996). Parameter estimation for linear dynamical systems. Retrieved from http://mlg.eng.cam.ac.uk/zoubin/papers/tr-96-2.pdf

Nguyen, H. T. T., & Galelli, S. (2018). A Linear Dynamical Systems Approach to Streamflow Reconstruction Reveals History of Regime Shifts in Northern Thailand. Water Resources Research, 54(3), 2057–2077. https://doi.org/10.1002/2017WR022114

Shumway, R. H., & Stoffer, D. S. (1982). An Approach to The Time Series Smoothing and Forecasting Using the EM Algorithm. Journal of Time Series Analysis, 3(4), 253–264. https://doi.org/10.1111/j.1467-9892.1982.tb00349.x

Tierney, J. E., Abram, N. J., Anchukaitis, K. J., Evans, M. N., Giry, C., Kilbourne, K. H., … Zinke, J. (2015). Tropical sea surface temperatures for the past four centuries reconstructed from coral archives. Paleoceanography, 30(3), 226–252. https://doi.org/10.1002/2014PA002717
